$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-05-04 src/s_cauchy.Rmd
added new rhat plot for eight schools

## 10,000 iterations
Nominal model with 50000 iterations and thinning by 5
```{r}
fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
            refresh = 0, iter = 50000, thin=5)
saveRDS(fit, "../output/fit_nominal_50000.rds")
# fit <- readRDS("../output/fit_nominal_50000.rds")
full_data <- rstan::extract(fit, permuted=F)

@@

Alternative model with 50000 iterations and thinning by 5
```{r}
fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
            refresh = 0, iter = 60000, thin=3)
saveRDS(fit, "../output/fit_alt_50000.rds")
full_data <- rstan::extract(fit, permuted=F)

@@

Compare quantiles
```{r}
fit <- readRDS("../output/fit_nominal_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x")
r_squared <- function(x, lower=0.001){
  qs <- seq(lower, 1-lower, lower)
  q_actual <- map_dbl(qs, ~qcauchy(.))
  r_vals <- vector(length = 50)
  for(i in seq_along(r_vals)){
    q_approx <- quantile(x[, , i], qs)
    aLM <- lm(data=tibble(q_approx, q_actual), formula = q_actual~q_approx)
    a_sum <- summary(aLM)
    r_vals[i] <- a_sum$r.squared
  }
  return(r_vals)
}

log_last <- log10(25000)
log_first <- log10(100)
iter <- round(10^(seq(log_first, log_last, length.out = 20)))
r_vals1 <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
r_vals1_lower <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.025))
r_vals1_upper <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.975))

fit <- readRDS("../output/fit_alt_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x")
r_vals <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
r_vals_lower <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.025))
r_vals_upper <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.975))
```

@@

@@

@@

Look at quantile accuracy for models converged by R* and Rhat
```{r}
fit <- readRDS("../output/fit_alt_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x")
quantile_errors <- function(x, lower=0.001){
  print(dim(x))
  qs <- seq(lower, 1-lower, lower)
  q_actual <- map_dbl(qs, ~qcauchy(.))
  errors <- matrix(nrow = length(qs),
                   ncol=dim(x)[3])
  for(i in 1:50){
    q_approx <- quantile(x[, , i], qs)
    errors[, i] <- abs(q_actual - q_approx)
  }
  errors <- errors %>% 
    as.data.frame() %>% 
    mutate(q=0.5-abs(0.5-qs)) %>% 
    melt(id.vars="q")
  return(errors)
}

errors_df <- quantile_errors(x, lower=0.001)
errors_df %>% 
  ggplot(aes(x=q, y=value)) +
  geom_jitter(width = 0.05) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method="lm") +
  xlab("Distance from centre") +
  ylab("Error")

aLM <- lm(data=errors_df, formula="log(value)~log(q)")
summary(aLM)

errors_df <- quantile_errors(x[1:iter[8], , ], lower=0.001)
errors_df %>% 
  ggplot(aes(x=q, y=value)) +
  geom_jitter(width = 0.05) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method="lm") +
  xlab("Distance from centre") +
  ylab("Error")

bLM <- lm(data=errors_df, formula="log(value)~log(q)")
summary(bLM)
```

>>>>

## 10,000 iterations
Nominal model with 50000 iterations
```{r}
fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
            refresh = 0, iter = 50000)
saveRDS(fit, "../output/fit_nominal_50000.rds")
fit <- readRDS("../output/fit_nominal_50000.rds")
full_data <- rstan::extract(fit, permuted=F)

@@

Alternative model with 50000 iterations
```{r}
fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
            refresh = 0, iter = 50000)
saveRDS(fit, "../output/fit_alt_50000.rds")
fit <- readRDS("../output/fit_alt_50000.rds")
full_data <- rstan::extract(fit, permuted=F)

@@

## Look at how various measures indicate convergence with iteration
Compare quantiles
```{r}
fit <- readRDS("../output/fit_nominal_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
r_squared <- function(x, lower=0.001){
  qs <- seq(lower, 1-lower, lower)
  q_actual <- map_dbl(qs, ~qcauchy(.))
  r_vals <- vector(length = 50)
  for(i in seq_along(r_vals)){
    q_approx <- quantile(x[, , i], qs)
    aLM <- lm(data=tibble(q_approx, q_actual), formula = q_actual~q_approx)
    a_sum <- summary(aLM)
    r_vals[i] <- a_sum$r.squared
  }
  return(r_vals)
}

log_last <- log10(10000)
log_first <- log10(100)
iter <- round(10^(seq(log_first, log_last, length.out = 20)))
r_vals1 <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
r_vals1_lower <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.025))
r_vals1_upper <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.975))

fit <- readRDS("../output/fit_alt_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
r_vals <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
r_vals_lower <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.025))
r_vals_upper <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.975))
```

@@

Cramer Von Mises -- seems very noisy
```{r}
library(goftest)
f_cvm <- function(x){
  temp <- cvm.test(x, null = "pcauchy")
  return(temp$p.value)
}

f_cvm_all_dims <- function(x, max_iter){
  vals <- map_dbl(seq(1, 50, 1), ~f_cvm(x[1:max_iter, , .]))
  return(vals)
}

fit <- readRDS("../output/fit_nominal_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
cvm_min <- map_dbl(iter, ~mean(f_cvm_all_dims(x, .)))

fit <- readRDS("../output/fit_alt_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
cvm_min <- map_dbl(iter, ~mean(f_cvm_all_dims(x, .)))
```

Anderson-Darling test -- again, noisy
```{r}
library(goftest)
f_ad <- function(x){
  temp <- ad.test(x, null = "pcauchy")
  return(temp$p.value)
}

f_ad_all_dims <- function(x, max_iter){
  vals <- map_dbl(seq(1, 50, 1), ~f_ad(x[1:max_iter, , .]))
  return(vals)
}

fit <- readRDS("../output/fit_nominal_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
cvm_min <- map_dbl(iter, ~mean(f_ad_all_dims(x, .)<0.05))

fit <- readRDS("../output/fit_alt_50000.rds")
x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
cvm_min <- map_dbl(iter, ~mean(f_ad_all_dims(x, .)<0.05))
```

@@

Progression of R* with ranked samples instead
```{r}

```

@@

### 2020-05-04 src/s_eight_schools.Rmd
added new rhat plot for eight schools

## Break each chain into two then make flattened data
Centered
```{r}
x <- rstan::extract(fit_cp, permuted=F)

r_star_centered <- r_star_dist(x)
```

Noncentered
```{r}
x <- rstan::extract(fit_ncp, permuted=F)
r_star_noncentered <- r_star_dist(x)
```

Store both
```{r}
c_df <- tibble(centered=r_star_centered,
               non_centered=r_star_noncentered)
```

## Trying with original data (i.e. 4 chains)
```{r}
x <- rstan::extract(fit_cp, permuted=F)
r_star_centered <- r_star_dist(x)

# noncentered
x <- rstan::extract(fit_ncp, permuted=F)
r_star_noncentered <- r_star_dist(x)
```

```{r}
d_df <- tibble(centered=r_star_centered,
               non_centered=r_star_noncentered)
```

Plot two R* distributions
```{r}
g1 <- 
  c_df %>% 
  melt(id.vars=NULL) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8) +
  xlim(0.78, 1.6) +
  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  ggtitle("A.") +
  theme(text = element_text(size=14, colour="black"),
        legend.position = "none")
g2 <- 
  d_df %>% 
  melt(id.vars=NULL) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8) +
  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
  xlim(0.78, 1.6) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  ggtitle("B.") +
  theme(text = element_text(size=14, colour="black"))
# a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")

# pdf("../output/eight_schools.pdf", width = 12, height = 6)
# multiplot(g1, g2, cols = 2)
# dev.off()
```

>>>>

## Break each chain into two then make flattened data
```{r}
# Centered
x <- rstan::extract(fit_cp, permuted=F)
r_star_centered <- r_star_dist(x)
mon_cp <- monitor(split_data(x))

# Noncentered
x <- rstan::extract(fit_ncp, permuted=F)
r_star_noncentered <- r_star_dist(x)
mon_ncp <- monitor(split_data(x))

c_df <- tibble(centered=r_star_centered,
               non_centered=r_star_noncentered,
               metric="R*",
               simulations="split")
c1_df <- tibble(centered=mon_cp$Rhat[1:10],
                non_centered=mon_ncp$Rhat[c(1, 2, 11:18)],
                metric="split-Rhat",
               simulations="split")
```

## Trying with original data (i.e. 4 chains)
```{r}
x <- rstan::extract(fit_cp, permuted=F)
r_star_centered <- r_star_dist(x, split_chains = F)
mon_cp <- monitor(fit_cp)
# noncentered
x <- rstan::extract(fit_ncp, permuted=F)
r_star_noncentered <- r_star_dist(x, split_chains = F)
mon_ncp <- monitor(fit_ncp)

d_df <- tibble(centered=r_star_centered,
               non_centered=r_star_noncentered,
               metric="R*",
               simulations="unsplit")
d1_df <- tibble(centered=mon_cp$Rhat[1:10],
                non_centered=mon_ncp$Rhat[c(1, 2, 11:18)],
                metric="split-Rhat",
               simulations="unsplit")
```

Plot two R* distributions
```{r}
g1 <- 
  c_df %>% 
  melt(id.vars=c("metric", "simulations")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8) +
  xlim(0.78, 1.6) +
  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  ggtitle("A.") +
  theme(text = element_text(size=14, colour="black"),
        legend.position = "none")
g2 <- 
  d_df %>% 
  melt(id.vars=c("metric", "simulations")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8) +
  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
  xlim(0.78, 1.6) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  ggtitle("B.") +
  theme(text = element_text(size=14, colour="black"),
        legend.position = "none")
# a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
```

Look at Rhat distributions
```{r}
e_df <- c1_df %>% 
  mutate(var_num=seq_along(metric)) %>% 
  bind_rows(d1_df %>% mutate(var_num=seq_along(metric)))
g3 <- e_df %>%
  select(-metric) %>% 
  melt(id.vars=c("simulations", "var_num")) %>% 
  pivot_wider(id_cols=c("variable", "var_num"), values_from = "value", names_from = "simulations") %>% 
  ggplot(aes(x=unsplit, y=split, colour=as.factor(variable))) +
  geom_point() +
  scale_x_continuous(limits=c(NA, 1.1)) +
  scale_y_continuous(limits=c(NA, 1.1)) +
  geom_vline(xintercept = 1.01, linetype=2) +
  geom_hline(yintercept = 1.01, linetype=2) +
  geom_abline(linetype=3) +
  scale_color_grey("Series", labels=c("centered", "non-centered")) +
  ylab("Split chains") +
  xlab("Original chains") +
  ggtitle("C.")

pdf("../output/eight_schools.pdf", width = 12, height = 4)
multiplot(g1, g2, g3, cols = 3)
dev.off()
```

### 2020-05-04 write_up/ms.tex
added new rhat plot for eight schools

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains. In all cases, the plots show 1000 $R^*$ draws. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:eight_schools}
\end{figure}

>>>>

$\widehat{R}$, like $R^*$, is recommended to be calculated using split chains. In Fig. \ref{fig:eight_schools}C, we plot the values of values of this metric when using the original four chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models. These show that the values of $\widehat{R}$ for the non-centered model for both the split and unsplit cases were, for all parameters, above 1.01, echoing the results for $R^*$; also, like $R^*$, the value of $\widehat{R}$ is higher for the split case. Finally, as for $R^*$, all parameters for the centered models were below 1.01, indicating convergence.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains vs those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line. In all cases, the plots show 1000 $R^*$ draws. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:eight_schools}
\end{figure}
